---
title: "COVID-19"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(lubridate)
```

## COVID-19 chaos model

Chaos model for COVID-19 european data.

### Get data

Get data from [European Center for Disease Control (ECDC)](https://www.ecdc.europa.eu).

```{r}
data <- read.csv("https://opendata.ecdc.europa.eu/covid19/casedistribution/csv", 
                 na.strings = "", 
                 fileEncoding = "UTF-8-BOM", 
                 stringsAsFactors = F)

# construct "date_reported" <date> column
data$date_reported <- mdy(paste0(data$month,"-",data$day,"-",data$year))
```

Isolate data from Europe, 
from countries or territories with big populations (remove islands & small states)
and select relevant columns only.

```{r}
eu <- data[data$continentExp == "Europe" & data$popData2019 >= 1000000,] %>%
  select("date_reported", "cases", "deaths", "countriesAndTerritories", "popData2019")
head(eu)
```

### Transform into timeseries

Get cases and deaths logdiffs per country.

```{r warning=FALSE}
isValidLogDiff <- function(x) {
  !is.nan(x) & is.finite(x) & x != 0
}

euLogDiff <- eu %>% 
  arrange(date_reported) %>%
  group_by(countriesAndTerritories) %>%
  mutate(
    cases_logdiff = (log(cases) - log(lag(cases))),
    deaths_logdiff = log(deaths) - log(lag(deaths))) %>%
  tail(-1) %>% filter(isValidLogDiff(cases_logdiff), isValidLogDiff(deaths_logdiff)) %>%
  group_split()

names(euLogDiff) = unique(eu$countriesAndTerritories)
```

### Plot Italy

Plot raw timeseries.

```{r}
ggplot(euLogDiff$Italy %>% 
         select(date_reported, cases, deaths) %>%
         pivot_longer(-date_reported, names_to = "variable", values_to = "value"),
       aes(date_reported, value, colour = variable)) +
  geom_point()
```

Plot logdiff.

```{r}
ggplot(euLogDiff$Italy %>% 
         select(date_reported, cases_logdiff, deaths_logdiff) %>%
         pivot_longer(-date_reported, names_to = "variable", values_to = "value"),
       aes(date_reported, value, colour = variable)) +
  geom_point()
```


